Text embedding models yield high-resolution insights into conceptual knowledge from short multiple-choice quizzes

### 2023-10-24 paper/main.tex
U with no tail

@@

If we are able to accurately estimate a participant's knowledge about the
content tested by a given question, our estimates of their knowledge should
carry some predictive information about whether the participant is likely to
answer that question correctly or incorrectly. We developed a statistical
approach to test this claim. For each question, in turn, we used
Equation~\ref{eqn:prop} to estimate each participant's knowledge at the given
question's embedding space coordinate, using all \textit{other} questions that
participant answered on the same quiz. For each quiz, we grouped these
estimates into two distributions: one for the estimated knowledge at the
coordinates of \textit{correctly} answered questions, and another for the
estimated knowledge at the coordinates of \textit{incorrectly} answered
questions (Fig.~\ref{fig:predictions}). We then used Mann-Whitney $U$-tests to
compare the means of these distributions of estimated knowledge.

@@

    \caption{\textbf{Estimating knowledge at the embedding coordinates of
    held-out questions.} Separately for each quiz (panel), we plot the
    distributions of predicted knowledge at the embedding coordinates of each
    held-out correctly (blue) or incorrectly (red) answered question. The
    Mann-Whitney $U$-tests reported in each panel are between the distributions
    of estimated knowledge at the coordinates of correctly versus incorrectly
    answered held-out questions. In the top row (``All questions''), we used
    all quiz questions (from each quiz, for each participant) except one to
    estimate knowledge at the held-out question's embedding coordinate. In the
    middle rows (``Across-lecture''), we used all questions about one lecture
    to estimate knowledge at the embedding coordinate of a held-out question
    about the \textit{other} lecture. In the bottom row (``Within-lecture''),
    we used all but one question about one lecture to estimate knowledge at the
    embedding coordinate of a held-out question about the \textit{same}
    lecture. We repeated each of these analyses using all possible held-out
    questions for each quiz and participant.}

@@

For the initial quizzes participants took (prior to watching either lecture),
participants' estimated knowledge tended to be low overall, and relatively
unstructured (Fig.~\ref{fig:predictions}, left column). When we held out
individual questions and estimated their knowledge at the held-out questions'
embedding coordinates, we found no reliable differences in the estimates when
the held-out question had been correctly versus incorrectly answered. This
``null'' effect persisted when we used \textit{all} of the Quiz 1 questions
from a given participant to predict a held-out question (``All questions''; $U
= 50587, p = 0.723$), when we used questions from one lecture to predict
knowledge at the embedding coordinate of a held-out question about the
\textit{other} lecture (``Across-lecture''; predicting knowledge for held-out
\textit{Four Fundamental Forces Questions} using \textit{Birth of Stars}
questions: $U = 8244, p = 0.184$; predicting knowledge for held-out
\textit{Birth of Stars} questions: $U = 8202.5, p = 0.161$), and when we used
questions from one lecture to predict knowledge at the embedding coordinate of
a held-out question about the \textit{same} lecture (``Within-lecture'';
\textit{Four Fundamental Forces}: $U = 7681.5, p = 0.746$; \textit{Birth of
Stars}: $U = 8125, p = 0.204$). We believe that this reflects a floor effect:
when knowledge is low everywhere, there is little signal to differentiate
between what is known versus unknown.

After watching \textit{Four Fundamental Forces}, estimated knowledge for
held-out correctly answered questions (from the second quiz;
Fig.~\ref{fig:predictions}, middle column) exhibited a positive shift relative
to held-out incorrectly answered questions. This held when we included all
questions in the analysis ($U = 58332, p < 0.001$), when we predicted knowledge
across-lectures (\textit{Four Fundamental Forces}: $U = 6749.5, p = 0.014$;
\textit{Birth of Stars}: $U = 8480, p = 0.016$), and when we predicted
knowledge at the embedding coordinates of held-out \textit{Four Fundamental
Forces} questions using other \textit{Four Fundamental Forces} questions from
the same quiz and participant ($U = 7224, p < 0.001$). This difference did
\textit{not} hold for within-lecture predictions of \textit{Birth of Stars}
knowledge ($U = 7419, p = 0.739$). Again, we suggest that this might reflect a
floor effect whereby knowledge about the content of the \textit{Birth of Stars}
material is relatively low everywhere in that region of text embedding space.

Finally, after watching \textit{Birth of Stars}, estimated knowledge for
held-out correctly answered questions (from the third quiz;
Fig.~\ref{fig:predictions}, right column) was higher for held-out correctly
answered questions than for held-out incorrectly answered questions. This held
when we included all questions in the analysis ($U = 38279, p = 0.022$), when
we carried out across-lecture predictions (\textit{Four Fundamental Forces}: $U
= 6684.5, p = 0.032$; \textit{Birth of Stars}: $U = 6414.5, p = 0.002$), and
and when we carried out within-lecture predictions of held-out \textit{Birth of
Stars} questions using other \textit{Birth of Stars} questions from the same
quiz and participant ($U = 6126, p = 0.006$). However, we found the
\textit{opposite} effect when we carried out within-lecture predictions of
held-out \textit{Four Fundamental Forces} questions using other \textit{Four
Fundamental Forces} questions from the same quiz and participant ($U = 6734, p
= 0.027$). Specifically, held-out correctly answered Quiz 3 questions about
\textit{Four Fundamental Forces} had reliably \textit{lower} estimated
knowledge than held-out incorrectly answered questions. Speculatively, we
suggest that this may reflect participants forgetting some of the \textit{Four
Fundamental Forces} content. If this forgetting happens in a relatively
``random'' way (with respect to spatial distance within the text embedding
space), then it could explain why some held-out questions about \textit{Four
Fundamental Forces} were answered incorrectly, even if questions at nearby
coordinates (i.e., about similar content) were answered correctly. This might
lead our approach to over-estimate knowledge for held-out questions about
``forgotten'' knowledge that participants answered incorrectly.

>>>>

\newcommand{\U}{{\fontfamily{serif}\selectfont\ensuremath{\mathrm{U}}}}

@@

If we are able to accurately estimate a participant's knowledge about the
content tested by a given question, our estimates of their knowledge should
carry some predictive information about whether the participant is likely to
answer that question correctly or incorrectly. We developed a statistical
approach to test this claim. For each question, in turn, we used
Equation~\ref{eqn:prop} to estimate each participant's knowledge at the given
question's embedding space coordinate, using all \textit{other} questions that
participant answered on the same quiz. For each quiz, we grouped these
estimates into two distributions: one for the estimated knowledge at the
coordinates of \textit{correctly} answered questions, and another for the
estimated knowledge at the coordinates of \textit{incorrectly} answered
questions (Fig.~\ref{fig:predictions}). We then used Mann-Whitney $\U$-tests to
compare the means of these distributions of estimated knowledge.

@@

    \caption{\textbf{Estimating knowledge at the embedding coordinates of
    held-out questions.} Separately for each quiz (panel), we plot the
    distributions of predicted knowledge at the embedding coordinates of each
    held-out correctly (blue) or incorrectly (red) answered question. The
    Mann-Whitney $\U$-tests reported in each panel are between the distributions
    of estimated knowledge at the coordinates of correctly versus incorrectly
    answered held-out questions. In the top row (``All questions''), we used
    all quiz questions (from each quiz, for each participant) except one to
    estimate knowledge at the held-out question's embedding coordinate. In the
    middle rows (``Across-lecture''), we used all questions about one lecture
    to estimate knowledge at the embedding coordinate of a held-out question
    about the \textit{other} lecture. In the bottom row (``Within-lecture''),
    we used all but one question about one lecture to estimate knowledge at the
    embedding coordinate of a held-out question about the \textit{same}
    lecture. We repeated each of these analyses using all possible held-out
    questions for each quiz and participant.}

@@

For the initial quizzes participants took (prior to watching either lecture),
participants' estimated knowledge tended to be low overall, and relatively
unstructured (Fig.~\ref{fig:predictions}, left column). When we held out
individual questions and estimated their knowledge at the held-out questions'
embedding coordinates, we found no reliable differences in the estimates when
the held-out question had been correctly versus incorrectly answered. This
``null'' effect persisted when we used \textit{all} of the Quiz 1 questions
from a given participant to predict a held-out question (``All questions''; $\U
= 50587, p = 0.723$), when we used questions from one lecture to predict
knowledge at the embedding coordinate of a held-out question about the
\textit{other} lecture (``Across-lecture''; predicting knowledge for held-out
\textit{Four Fundamental Forces Questions} using \textit{Birth of Stars}
questions: $\U = 8244, p = 0.184$; predicting knowledge for held-out
\textit{Birth of Stars} questions: $\U = 8202.5, p = 0.161$), and when we used
questions from one lecture to predict knowledge at the embedding coordinate of
a held-out question about the \textit{same} lecture (``Within-lecture'';
\textit{Four Fundamental Forces}: $\U = 7681.5, p = 0.746$; \textit{Birth of
Stars}: $\U = 8125, p = 0.204$). We believe that this reflects a floor effect:
when knowledge is low everywhere, there is little signal to differentiate
between what is known versus unknown.

After watching \textit{Four Fundamental Forces}, estimated knowledge for
held-out correctly answered questions (from the second quiz;
Fig.~\ref{fig:predictions}, middle column) exhibited a positive shift relative
to held-out incorrectly answered questions. This held when we included all
questions in the analysis ($\U = 58332, p < 0.001$), when we predicted knowledge
across-lectures (\textit{Four Fundamental Forces}: $\U = 6749.5, p = 0.014$;
\textit{Birth of Stars}: $\U = 8480, p = 0.016$), and when we predicted
knowledge at the embedding coordinates of held-out \textit{Four Fundamental
Forces} questions using other \textit{Four Fundamental Forces} questions from
the same quiz and participant ($\U = 7224, p < 0.001$). This difference did
\textit{not} hold for within-lecture predictions of \textit{Birth of Stars}
knowledge ($\U = 7419, p = 0.739$). Again, we suggest that this might reflect a
floor effect whereby knowledge about the content of the \textit{Birth of Stars}
material is relatively low everywhere in that region of text embedding space.

Finally, after watching \textit{Birth of Stars}, estimated knowledge for
held-out correctly answered questions (from the third quiz;
Fig.~\ref{fig:predictions}, right column) was higher for held-out correctly
answered questions than for held-out incorrectly answered questions. This held
when we included all questions in the analysis ($\U = 38279, p = 0.022$), when
we carried out across-lecture predictions (\textit{Four Fundamental Forces}: $\U
= 6684.5, p = 0.032$; \textit{Birth of Stars}: $\U = 6414.5, p = 0.002$), and
and when we carried out within-lecture predictions of held-out \textit{Birth of
Stars} questions using other \textit{Birth of Stars} questions from the same
quiz and participant ($\U = 6126, p = 0.006$). However, we found the
\textit{opposite} effect when we carried out within-lecture predictions of
held-out \textit{Four Fundamental Forces} questions using other \textit{Four
Fundamental Forces} questions from the same quiz and participant ($\U = 6734, p
= 0.027$). Specifically, held-out correctly answered Quiz 3 questions about
\textit{Four Fundamental Forces} had reliably \textit{lower} estimated
knowledge than held-out incorrectly answered questions. Speculatively, we
suggest that this may reflect participants forgetting some of the \textit{Four
Fundamental Forces} content. If this forgetting happens in a relatively
``random'' way (with respect to spatial distance within the text embedding
space), then it could explain why some held-out questions about \textit{Four
Fundamental Forces} were answered incorrectly, even if questions at nearby
coordinates (i.e., about similar content) were answered correctly. This might
lead our approach to over-estimate knowledge for held-out questions about
``forgotten'' knowledge that participants answered incorrectly.
